PMC Survey Analysis Part 2

### 2021-12-07 analysis_code/analysis1.Rmd
tidy4

---
title: "PMC Survey QC Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.

@@

x <- read.table("data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
head(x)
colnames(x)
dim(x)

@@

## Score

Here I am proposing a scoring scheme.
If there is information missing from the article, or basic mistakes are made, then a point is deducted.
If the article goes over and above the basic reproducibility, then they are awarded a point.

Gene set library origin not stated = -1

Gene set library version not stated = -1

Stat test not stated = -1

No stat test conducted = -1

No FDR correction conducted = -1

App used not stated = -1

App version not stated = -1

Background list not defined = -1

Inappropriate background list used = -1

Code availability = +1

Gene lists provided = +1

```{r,score}

score <- function(r){
  r[is.na(r)] <- 0
  SCORE=0
  # gene set lib
  if(r["Gene.set.library"]=="Not stated"){
    SCORE=SCORE-1
  }
  # GS version
  if(r["GS.version"]=="No"){
    SCORE=SCORE-1
  }
  # stat test
  if(r["Statistical.test.used"]=="No test"){
    SCORE=SCORE-1
  }
  if(r["Statistical.test.used"]=="Not stated"){
    SCORE=SCORE-1
  }
  # FDR
  if(r["FDR.Correction"]!="Yes"){
    SCORE=SCORE-1
  }
  # app used
  if(r["App.used"]=="Not stated"){
    SCORE=SCORE-1
  }
  # App version
  if(r["App.Version"]=="No"){
    SCORE=SCORE-1
  }
  # Code availability
  if(r["Code.availability"]=="Yes"){
    SCORE=SCORE+1
  }
  # Background
  if(r["Background.gene.set"]=="Not stated"){
    SCORE=SCORE-1
  }
  if(r["Background.gene.set"]=="No"){
    SCORE=SCORE-1
  }
  if(r["Background.gene.set"]=="Stated, but incorrect"){
    SCORE=SCORE-1
  }
  # gene list
  if(r["Gene.lists.provided"]=="Yes"){
    SCORE=SCORE+1
  }
  return(SCORE)
}

scores <- apply(X = x,MARGIN = 1, FUN = score)
barplot(table(scores), xlab="analysis score",ylab="frequency")
x$scores <- scores
length(which(scores>0))

mean(x$scores)
sd(x$scores)
median(x$scores)

## Journal rank

```{r,journal rank}

jmetrics <- read.table("scimagojr_2020.csv",sep=";",header=TRUE,dec = ",")
jmetrics$Title <- toupper(jmetrics$Title)
head(jmetrics,3)
dim(jmetrics)

nlmcat <- readLines("nlmcatalog_result.txt")
journaltitle <- nlmcat[grep("Title\\(s\\):",nlmcat)]
journaltitle <- gsub("Title\\(s\\): ","",journaltitle)
journaltitle <- gsub("\\.$","",journaltitle)
journaltitle <- toupper(journaltitle) 
journalabbrev <- nlmcat[grep("Title Abbreviation:",nlmcat)]
journalabbrev <- sapply(strsplit(journalabbrev,":"),"[[",2)
journalabbrev <- gsub(" $","",journalabbrev)
journalabbrev <- gsub("^ ","",journalabbrev)
jdf <- data.frame(journalabbrev,journaltitle)

mjdf <- merge(jdf,jmetrics,by.x="journaltitle",by.y="Title")

xm <- merge(x,mjdf,by.x="Journal",by.y="journalabbrev",all.x = TRUE)

## Analysis scores vs Journal rank

```{r,scoresvsjrank1}

tail(xm$Journal)

mylm1 <- lm (xm$scores ~ xm$SJR)
plot(xm$SJR,xm$scores, bty="n",
  xlab="SJR",ylab="score",main="Association of analysis scores with journal metrics")
abline(mylm1,col="red")

cor.test(xm$scores,xm$SJR,method="pearson")

cor.test(xm$scores,xm$SJR,method="spearman")

## Best and worst journals

Minimum of 5 analyses.

```{r,goodbadugly}

xmm <-xm[,c("Journal","scores")]
tab <- table(xmm$Journal)
tab <- tab[which(tab>3)]
xmm <- xmm[which(xmm$Journal %in% names(tab)),]
dim(xmm)
jres <- aggregate(. ~ Journal, xmm ,mean)
jsd <- aggregate(. ~ Journal, xmm ,sd)
jres$sd <- jsd$scores
jres <- jres[order(jres$scores),]
dim(jres)
head(jres)

par(mar=c(5,12,3,1))
barplot(tail(jres$scores,20),names.arg = tail(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
        main = "Highest scoring journals",xlim=c(-4,0))
grid()

par(mar=c(5,12,3,1))
barplot(head(jres$scores,20),names.arg = head(jres$Journal,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="mean score",
        main = "Lowest scoring journals",xlim=c(-5,0))
grid()

j <- unique(xmm$Journal)
jscores <- sapply(j,function(jj) {
 xmm[which(xmm$Journal == jj ),2]  
})

jscores <- jscores[order(unlist(lapply(jscores,mean)))]

par(mar=c(5,12,3,1))
vioplot(tail(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Highest scoring journals")

par(mar=c(5,12,3,1))
vioplot(head(jscores,20),horizontal = TRUE,las=1,cex.axis=0.75,main="Lowest scoring journals")

>>>>

---
title: "PMC Survey Analysis Part 1"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

Here we are performing an analysis of 200 articles which is randomly selected from 1500 PMC articles.
These articles were examined independently by two team members and the results were compared.
Any inconsistencies were resolved.

The code shown here was used to generate Figure 2 of the manuscript.

@@

x <- read.table("../data/QC-analysis2.tsv",header=TRUE,fill=TRUE,sep="\t")
head(x)
colnames(x)
dim(x)

@@

### 2021-12-07 analysis_code/analysis2.Rmd
tidy4

---
title: "PMC Survey Data Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

Here we are performing an analysis of the ways enrichment analysis has been done in a sample of 1500 PMC articles from 2019.

@@

## Journal

```{r,Journal1}

journal <- x$Journal
journal_split <- strsplit(journal,", ")
journal <- unlist(journal_split)
res <- table(journal)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
#names(res) <- gsub("Gene expression array","RNA array",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Journal", xlim=c(0,100))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Journal", xlim=c(0,1200))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
grid()

dir.create("images")
png("images/journals1.png")
par(mar=c(5,8,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Journal", xlim=c(0,1200))
text(y = (1:length(res2)*1.2)-0.8  , x = res2+50, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()
```

## Omics type

```{r,omics1}

omics <- x$Omics.type
omics_split <- strsplit(omics,", ")
omics <- unlist(omics_split)
res <- table(omics)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Gene expression array","RNA array",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

names(res) <- gsub("RNA array","Gene expression array",names(res))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Omics type", xlim=c(0,600))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Omics type", xlim=c(0,600))
text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
grid()

png("images/omics1.png")
par(mar=c(5,10,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Omics type", xlim=c(0,600))
text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## Organism

```{r,org1}

org <- x$Organism
org_split <- strsplit(org,", ")
org <- unlist(org_split)
res <- table(org)
res <- res[order(res)]
length(res)
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"), scale=c(4,.5))

par(mar=c(5,12,3,1))
names(res) <- gsub("human","Homo sapiens",names(res))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Organism", xlim=c(0,1200))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Organism", xlim=c(0,1300))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()

png("images/organisms1.png")
par(mar=c(5,10,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Organism", xlim=c(0,1300))
text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## Gene set library

```{r,gsl1}

GSL <-x$Gene.set.library
GSL_split <- strsplit(GSL,", ")
GSL <- unlist(GSL_split)
res <- table(GSL)
res <- res[order(res)]
length(res)
which(names(res)=="Not stated")/sum(res)*100
res

par(mar=c(1,1,1,1))
names(res) <- gsub("Homo sapiens","human",names(res))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,1000))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,1100))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()

png("images/genesetlib1.png")
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set library", xlim=c(0,1100))
text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()
```

## Gene set version

```{r,gsv1}

GSV <-x$GS.version
res <- table(GSV)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,1600))
text(y = (1:length(res)*1.2) - 0.6 , x = res-65, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/genesetvers1.png",width = 260, height = 260)
par(mar=c(5,7,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene set version defined", xlim=c(0,1900))
text(y = (1:length(res)*1.2) - 0.6 , x = res+200, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()
```

## Statistical test used

```{r,test1}

test <-x$Statistical.test.used
test <- strsplit(test,", ")
test <- unlist(test)
res <- table(test)
res <- res[order(res)]
res[which(names(res)=="Not stated")] / sum(res) * 100
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Test used", xlim=c(0,1200))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "Test used", xlim=c(0,1200))
text(y = (1:length(res2)*1.2)-1.2  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()

png("images/stattest1.png")
par(mar=c(5,10,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Test used", xlim=c(0,1200))
text(y = (1:length(res2)*1.2)-0.8  , x = res2+60, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()
```

## FDR Correction

```{r,fdr1}

fdr <-x$FDR.Correction
fdr <- strsplit(fdr,", ")
fdr <- unlist(fdr)
res <- table(fdr)
res <- res[order(res)]
sum(res[which(names(res)!="Yes")])/sum(res)*100
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR", xlim=c(0,900))
text(y = (1:length(res)*1.2)-0.7  , x = res-30, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/fdr1.png",width = 260, height = 260)
par(mar=c(5,7,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "FDR", xlim=c(0,1000))
text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## App used

```{r,app1}

App <-x$App.used
App_split <- strsplit(App,", ")
App <- unlist(App_split)
res <- table(App)
res <- res[order(res)]
res[which(names(res)=="Not stated")]/sum(res)*100
length(unique(subset(x,App.used=="Not stated")$Journal))
unique(subset(x,App.used=="Not stated")$Journal)
length(res)
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "App used", xlim=c(0,400))
grid()

other <- sum(res[1:(nrow(res)-10)])
res2 <- c(other,tail(res,9))
names(res2)[1] <- "Other"
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 0.7, xlab="no. analyses",
        main = "App used", xlim=c(0,500))
text(y = (1:length(res2)*1.2)-1.2  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
grid()

png("images/app1.png")
par(mar=c(5,12,3,1))
barplot(res2,horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App used", xlim=c(0,500))
text(y = (1:length(res2)*1.2)-0.8  , x = res2-20, label = res2, pos = 3, cex = 1, col = "black")
grid()
dev.off()

length(unique(subset(x,App.used=="Not stated")$Journal))

## App version

```{r,appv1}

APV <-x$App.Version
res <- table(APV)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,1600))
text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/appvers1.png",width = 260, height = 260)
par(mar=c(5,7,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "App version defined", xlim=c(0,1600))
text(y = (1:length(res)*1.2) - 0.6 , x = res-200, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## Code available

```{r,code1}

code <-x$Code.availability
res <- table(code)
res
res[1]/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,330))
text(y = (1:length(res)*1.2) - 0.6 , x = res+15, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/code1.png",width = 260, height = 260)
par(mar=c(5,7,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Code availability", xlim=c(0,350))
text(y = (1:length(res)*1.2) - 0.6 , x = res+30, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## Background gene set

```{r,background1}

BG <-x$Background.gene.set
res <- table(BG)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list defined", xlim=c(0,1300))
text(y = (1:length(res)*1.2)-0.7  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/bg1.png")
par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Background list defined", xlim=c(0,1400))
text(y = (1:length(res)*1.2)-0.7  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()
```

## Gene lists provided

```{r,genelists1}

GL <-x$Gene.lists.provided
res <- table(GL)
res
sum(res[which(names(res)!="Yes")])/sum(res)*100

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,1200))
text(y = (1:length(res)*1.2) - 0.6 , x = res-70, label = res, pos = 3, cex = 1, col = "black")
grid()
colnames(x)

png("images/genelists1.png",width = 260, height = 260)
par(mar=c(5,7,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Gene lists provided", xlim=c(0,1200))
text(y = (1:length(res)*1.2) - 0.6 , x = res-120, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()

## Assumptions violated

```{r,assumptions1}

ok <- nrow(subset(x,Assumptions.violated=="No"))
ok
bad <- nrow(subset(x,Assumptions.violated!="No"))
bad
ok/sum(bad,ok)*100

ass <-x$Assumptions.violated
ass <- strsplit(ass,", ")
ass <- unlist(ass)
res <- table(ass)
res <- res[order(res)]
res

par(mar=c(1,1,1,1))
wordcloud(words = names(res), freq = res, min.freq = 1, 
    max.words=200, random.order=FALSE, rot.per=0.35, 
    colors=brewer.pal(8, "Dark2"))

par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Assumptions violated", xlim=c(0,1500))
text(y = (1:length(res)*1.2)-0.75  , x = res+60, label = res, pos = 3, cex = 1, col = "black")
grid()

png("images/assumptions1.png")
par(mar=c(5,12,3,1))
barplot(tail(res,20),horiz=TRUE,las=1,cex.names = 1, xlab="no. analyses",
        main = "Assumptions violated", xlim=c(0,1500))
text(y = (1:length(res)*1.2)-0.75  , x = res+80, label = res, pos = 3, cex = 1, col = "black")
grid()
dev.off()

>>>>

---
title: "PMC Survey Analysis Part 2"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

Here we are performing the analysis that is described in Figure 3 of the manuscript.
This involves calculating an analysis score for a 1500 articles from 2019.
We then identify the best/worst journals, and correlate analysis scores with journal and article level metrics.

@@
